---
title: "spatial_fragmentation_measure"
output:
  pdf_document: default
  html_document: default
date: "2026-01-25"
---

### Overview

This .Rmd file goes through the procedure in steps for creating a new type of measure for spatial fragmentation based upon the distorted shapes observed in American municipal boundaries. Typically, research relies on simple aggregates weighted by population (e.g. number of municipalities per person). I construct a population HHI for each CBSA as a means of more accurately assessing the degree to which municipal authority (simply taxpayers) is concentrated or distributed, though introducing population into the fragmentation index itself may introduce endogeneity concerns through the potential for population sorting being correlated with that of firms. As such, I aim to create a measure that explicity uses the shapes of incorporated places as a means of analyzing fragmentation.
The strange shapes of American municipalities have been noted in publications such as The Economist (https://www.economist.com/united-states/2018/12/22/why-american-cities-are-so-weirdly-shaped). I aim to quantify this "strangeness," calculating a municipality's departure from a "normal" or "ideal" shape, which I code here as a minimum bounding circle or rectangle. The intuition is that these departures proxy the fragmentation of a CBSA as municipal competition results in the splintering of borders. This can be represented as a weighted average of percent of the ideal shape occupied across the entire CBSA, or simply this deviation for the central city (for instance, compare the borders of Seattle vs. Los Angeles), which serves as a less resource intensive measure of centrality. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
options(scipen = 0, digits = 5)  # controls base R output`
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, plotly, tidycensus, readxl, httr, jsonlite, glue, sf, mapview, tigris, crsuggest, lwgeom, spatstat, patchwork, rnaturalearth, showtext, systemfonts, kableExtra) 
```

Optional function in the case that unzipping geographic files is necessary

```{r}
# functions 
unzip_shape <- function(file_url, sf_name) {
  url <- glue(file_url)
  
  tmp <- tempfile(fileext = ".zip")
  download.file(url, tmp, mode = "wb")
  
  st_read(paste0("/vsizip/", tmp)) %>% 
    st_transform("EPSG:4326")
}
```

### Step-by-step procedure

#### Set-up

First, I retrieve a county-CBSA crosswalk file as well as a national shapefile for all counties in the US. These are joined to create a map of all CBSAs with their associated counties. Additionally, I create a water poly

```{r}
# cross-walk data to connect county and cbsa

xwalk.cbsa_county <- read_csv("https://data.nber.org/cbsa-csa-fips-county-crosswalk/2023/cbsa2fipsxw_2023.csv") %>% 
  mutate(
    fullfips = paste0(fipsstatecode,fipscountycode),
    cbsacode = as.character(cbsacode)
  ) %>% 
  filter(metropolitanmicropolitanstatis == "Metropolitan Statistical Area") %>% 
  select(cbsacode,cbsatitle,fipsstatecode,fipscountycode,fullfips)

# get county shapefiles (not included in tiger place shapefiles)

sf.counties <- counties(
  state = NULL,
  cb = TRUE,
) %>% 
  left_join(xwalk.cbsa_county, by = join_by(GEOID == fullfips))

# plot to ensure data loaded correctly

ggplot(sf.counties %>% 
         filter(!STATEFP %in% c("02","15")) %>% 
         filter(as.numeric(STATEFP) < 57)
       , aes(fill = cbsacode)) +
  geom_sf() +
  theme(legend.position = "none")

# Load the complete national place file. These are already clipped by water boundary, so no need to perform erase_water() yet.

sf.nat_places <- places(
  state = NULL,
  cb = TRUE
)

# Create a basic water filter using coastline and Great Lakes

sf_use_s2(FALSE)
sf.water_mask <- ne_download(scale = 10, type = "lakes", category = "physical", returnclass = "sf") %>% 
  filter(name %in% c("Lake Superior", "Lake Michigan", "Lake Huron", 
                     "Lake Erie", "Lake Ontario")) %>% 
  bind_rows(
    ne_download(scale = 10, type = "ocean", category = "physical", returnclass = "sf")
  ) %>% 
  st_make_valid() %>% 
  st_union() %>% 
  st_make_valid()
sf_use_s2(TRUE)
```
#### Main body

Next, I select a CBSA (in this case, Seattle), and find the requisite counties and TIGER place shapefiles, which are merged to create a full place shapefile for the CBSA.

```{r}
# initialize with CBSA code

needed_cbsacode = "42660"

needed_fips <- xwalk.cbsa_county %>% 
  filter(grepl(needed_cbsacode,cbsacode)) %>% 
  distinct(fullfips) %>% 
  pull(fullfips)

# Since places can span multiple counties, there is no identifier in the data. Thus, need to use a spatial clip

sf.places <-
  st_intersection(
    sf.nat_places %>% 
      filter(STATEFP %in% c(substr(needed_fips, start = 1, stop = 2) %>% 
                       unique())),
    sf.counties %>% 
      filter(GEOID %in% c(needed_fips)) %>% 
      select()
  ) %>% 
  filter(LSAD %in% c("25","43"))

needed_crs <- suggest_crs(sf.places) %>% 
  filter(crs_units == "m") %>% 
  pull(crs_code) %>% 
  first()

# occasionally, places will be duplicated, so need to filter out duplicates

sf.places <- st_transform(sf.places, as.numeric(needed_crs)) %>% 
  group_by(GEOID) %>%
  summarise(across(-geometry, first),
    geometry = st_union(geometry), .groups = "drop")

# check sf with plot

ggplot(sf.places) + 
  geom_sf()
```
With each place polygon, I now calculate the minimum bounding circle and rectangle as "ideal" shapes that would occur in theoretical equilibrium. For each municipality, a difference is then calculated as the percentage of this ideal shape filled by the actual municipal borders.

```{r}
# now calculate a spatial index

sf.buffers <- rbind(
  st_minimum_bounding_circle(sf.places) %>% 
    mutate(type = "area_circ"),
  st_minimum_rotated_rectangle(sf.places) %>% 
    mutate(type = "area_rect")
) %>% 
  select(GEOID,type) %>% 
  st_difference(
    sf.water_mask %>% st_transform(as.numeric(needed_crs))
  ) %>% 
  mutate(buffer_area = as.numeric(st_area(geometry))) %>% 
  st_drop_geometry() %>% 
  pivot_wider(names_from = type, values_from = buffer_area)

sf.places <- inner_join(
  sf.places, sf.buffers, by = join_by(GEOID == GEOID)
)

sf.places <- sf.places %>% st_difference(
    sf.water_mask %>% st_transform(as.numeric(needed_crs))
  ) %>% 
  mutate(re_area = as.numeric(st_area(geometry)),
         diff_circ = re_area/area_circ,
         diff_rect = re_area/area_rect) %>% 
  filter(re_area > 1000) %>% # this filters out extremely small areas (happens if clip performs weirdly)
  arrange(-diff_rect)
```
Below are some further visualizations.

```{r}
# visualize some of this data

p1 = ggplot() + 
  geom_histogram(data = sf.places %>% 
              st_drop_geometry(),
            aes(x = diff_circ),
            bins = 20)

p2 = ggplot() + 
  geom_histogram(data = sf.places %>% 
              st_drop_geometry(),
            aes(x = diff_rect),
            bins = 20)

p1 + p2

p3 = ggplot() +
  geom_sf(data = sf.places, aes(fill = diff_rect))

p4 = ggplot() +
  geom_sf(data = sf.places, aes(fill = diff_circ))

p3 + p4
```

The final step is converting these municipal-level figures into a global index for the city that can be used as a point of comparison. This is done through a weighted sum of each of the differences, where the weight is the percentage of total municipal land area occupied by the municipality.

```{r}
sf.places <- sf.places %>% 
  mutate(pct_area = re_area/sum(re_area))

index_circ <- sum((1-sf.places$diff_circ) * sf.places$pct_area)
index_rect <- sum((1-sf.places$diff_rect) * sf.places$pct_area)
```

### Function

This requires the set-up chunk to be performed - load in the national places and counties shapefiles as well as the crosswalk. The following chunk sets up 3 functions that will be used to construct the spatial fragmentation indices: the primary function is get_spat_diff_index, which repeats the steps outlined above; plot_places returns a plot of a specified CBSA to eye-test any areas of interest and safe_get_spat_diff_index wraps the main function to ensure errors do not disrupt assembly.

```{r}
get_spat_diff_index <- function(cbsacode) {
  needed_cbsacode <- as.character(cbsacode)
  
  needed_fips <- xwalk.cbsa_county %>% 
    filter(grepl(needed_cbsacode,cbsacode)) %>% 
    distinct(fullfips) %>% 
    pull(fullfips)
  
  sf.places <-
  st_intersection(
    sf.nat_places %>% 
      filter(STATEFP %in% c(substr(needed_fips, start = 1, stop = 2) %>% 
                       unique())),
    sf.counties %>% 
      filter(GEOID %in% c(needed_fips)) %>% 
      select()
  ) %>% 
  filter(LSAD %in% c("25","43")) %>% 
    st_make_valid()
  
  needed_crs <- suggest_crs(sf.places) %>% 
  filter(crs_units == "m") %>% 
  pull(crs_code) %>% 
  first()
  
  sf.places <- st_transform(sf.places, as.numeric(needed_crs)) %>% 
  group_by(GEOID) %>%
  summarise(across(-geometry, first),
    geometry = st_union(geometry), .groups = "drop")
  
  sf.buffers <- rbind(
  st_minimum_bounding_circle(sf.places) %>% 
    mutate(type = "area_circ"),
  st_minimum_rotated_rectangle(sf.places) %>% 
    mutate(type = "area_rect")
) %>% 
  select(GEOID,type) %>% 
    suppressMessages %>% 
    st_make_valid() %>% 
  st_difference(
    sf.water_mask %>% st_transform(as.numeric(needed_crs)) %>% 
    st_make_valid()
  ) %>% 
  mutate(buffer_area = as.numeric(st_area(geometry))) %>% 
  st_drop_geometry() %>% 
  pivot_wider(names_from = type, values_from = buffer_area)
  
  sf.places <- inner_join(
  sf.places, sf.buffers, by = join_by(GEOID == GEOID)
)

  sf.places <- sf.places %>% st_difference(
    sf.water_mask %>% st_transform(as.numeric(needed_crs))
  ) %>%
  mutate(re_area = as.numeric(st_area(geometry)),
         diff_circ = re_area/area_circ,
         diff_rect = re_area/area_rect) %>% 
  filter(re_area > 1000)
  
  sf.places <- sf.places %>% 
  mutate(pct_area = re_area/sum(re_area))

index_circ <- sum((1-sf.places$diff_circ) * sf.places$pct_area)
index_rect <- sum((1-sf.places$diff_rect) * sf.places$pct_area)

return(tibble(index_circ = index_circ,
              index_rect = index_rect))
}

plot_places <- function(cbsacode) {
  needed_cbsacode <- as.character(cbsacode)
  
  needed_fips <- xwalk.cbsa_county %>% 
    filter(grepl(needed_cbsacode,cbsacode)) %>% 
    distinct(fullfips) %>% 
    pull(fullfips)
  
  sf.places <-
  st_intersection(
    sf.nat_places %>% 
      filter(STATEFP %in% c(substr(needed_fips, start = 1, stop = 2) %>% 
                       unique())),
    sf.counties %>% 
      filter(GEOID %in% c(needed_fips)) %>% 
      select()
  ) %>% 
  filter(LSAD %in% c("25","43"))
  
  needed_crs <- suggest_crs(sf.places) %>% 
  filter(crs_units == "m") %>% 
  pull(crs_code) %>% 
  first()
  
  sf.places <- st_transform(sf.places, as.numeric(needed_crs)) %>% 
  group_by(GEOID) %>%
  summarise(across(-geometry, first),
    geometry = st_union(geometry), .groups = "drop")
  
  return(sf.places)
}

safe_get_spat_diff_index <- possibly(
  get_spat_diff_index, 
  otherwise = tibble(index_circ = NA_real_, index_rect = NA_real_)
)
```

Sanity check: test with the MSAs of Washington State.

```{r}
test_cbsas <- xwalk.cbsa_county %>% 
  filter(fipsstatecode == 53) %>% 
  distinct(cbsacode)

test_cbsas <- test_cbsas %>% 
  mutate(output = map(cbsacode, get_spat_diff_index, .progress = TRUE)) %>% 
  unnest(output)

# after testing, can check highest and lowest scores:
ggplot(data = plot.places())
```

## Top 100 MSAs

I now begin the formal assembly of CBSA scores. Here, only the top 100 MSAs are calculated, though this procedure can easily be expanded to all MSAs or even the ~900 CBSAs listed by the US Census.

```{r}
# get the top 100 MSAs by population
df.frag_index <- read_csv("data/fragmentation_basic.csv") %>% 
  arrange(-pop_total) %>% 
  slice(1:100)

# perform the spatial difference index function
df.frag_index <- df.frag_index %>% 
  mutate(output = map(cbsacode, safe_get_spat_diff_index, .progress = TRUE)) %>% 
  unnest(output)

# plot similarity to other measures of fragmentation
ggplot(data = df.frag_index, aes(x = hhi_pop_total, y = index_rect)) +
  geom_point() +
  geom_smooth(method = "lm")

write.csv(df.frag_index, file = "data/fragmentation_top_100.csv", row.names = FALSE)
```

I also perform an analysis that centers on the "central cities" of each CBSA (as defined by Census naming conventions, though this can be restructured to just include the largest city in each CBSA).

```{r}
df.central <- df.frag_index %>%
  mutate(cities = str_extract(cbsatitle, "^[^,]+"),
         cbsacode = as.character(cbsacode)) %>% 
  select(cbsacode,cities) %>% 
  separate_longer_delim(cities, "-") %>% 
  filter(str_trim(cities) != "") %>% 
  left_join(xwalk.cbsa_county %>% 
  select(cbsacode,fipsstatecode) %>% 
  unique()
  , by = "cbsacode", relationship = "many-to-many")

sf.central <- inner_join(
  sf.nat_places, df.central, by = join_by(NAME == cities, STATEFP == fipsstatecode)
) %>% 
  group_by(cbsacode, NAME) %>%
  slice_max(st_area(geometry), n = 1, with_ties = FALSE) %>%
  ungroup()

# Redundant code from before, don't need to call function because of limited total places

needed_crs <- suggest_crs(sf.central) %>% 
  filter(crs_units == "m") %>% 
  pull(crs_code) %>% 
  first()

sf.central <- st_transform(sf.central, as.numeric(needed_crs)) %>% 
  group_by(GEOID) %>%
  summarise(across(-geometry, first),
    geometry = st_union(geometry), .groups = "drop")

sf.central_buffers <- rbind(
  st_minimum_bounding_circle(sf.central) %>% 
    mutate(type = "area_circ"),
  st_minimum_rotated_rectangle(sf.central) %>% 
    mutate(type = "area_rect")
) %>% 
  select(GEOID,type) %>% 
  st_difference(
    sf.water_mask %>% st_transform(as.numeric(needed_crs))
  ) %>% 
  mutate(buffer_area = as.numeric(st_area(geometry))) %>% 
  st_drop_geometry() %>% 
  pivot_wider(names_from = type, values_from = buffer_area)
sf.central <- inner_join(
  sf.central, sf.central_buffers, by = join_by(GEOID == GEOID)
)
sf.central <- sf.central %>% st_difference(
    sf.water_mask %>% st_transform(as.numeric(needed_crs))
  ) %>% 
  mutate(re_area = as.numeric(st_area(geometry)),
         diff_circ = re_area/area_circ,
         diff_rect = re_area/area_rect)

# Simple average, assume that each listed city has equal weight in the metro

df.central_index <- sf.central %>% 
  st_drop_geometry() %>% 
  group_by(cbsacode) %>% 
  summarise(index_rect_central = mean(diff_rect),
            index_circ_central = mean(diff_circ)) %>% 
  inner_join(xwalk.cbsa_county %>% 
               select(cbsacode) %>% unique(),
             by = "cbsacode") 

df.frag_index = read_csv("data/fragmentation_top_100.csv") %>% 
  select(-cities) %>% 
  mutate(cbsacode = as.character(cbsacode)) %>% 
  left_join(df.central_index, by = "cbsacode")

write.csv(df.frag_index, file = "data/fragmentation_top_100.csv", row.names = FALSE)
```

## Figures

These chunks cover code written for the export of usable figures in the final paper. The first creates a plot for the Birmingham MSA, with the Birmingham municipality highlighted in red.

```{r}
sf.bir <- plot_places(13820) %>% 
  mutate(highlight = if_else(NAME == "Birmingham",1,0),
         highlight = as.character(highlight))

p.bir <- ggplot(data = sf.bir, aes(fill = highlight)) +
  theme_void() + 
  scale_fill_manual(values = c("1" = "red","0" = "lightgrey")) +
  geom_sf() +
  theme(legend.position = "none")

p.bir
ggsave("figures/bir_msa.png", plot = p.bir)
```
Next, I create visualizations for the ideal shape buffer process

```{r}
sf.bir <- sf.bir %>% 
  filter(NAME == "Birmingham") %>% 
  mutate(area = st_area(geometry))

sf.bir_circ <- st_minimum_bounding_circle(sf.bir) %>% 
  mutate(area = st_area(geometry))
sf.bir_rect <- st_minimum_rotated_rectangle(sf.bir) %>% 
  mutate(area = st_area(geometry))

systemfonts::system_fonts() %>% 
  filter(grepl("Times", name, ignore.case = TRUE))

sf.bir$area[1]/sf.bir_circ$area[1]

p.bir_circ <- ggplot() +
  geom_sf(data = sf.bir_circ, fill = "white") +
  geom_sf(data = sf.bir, fill = "red") +
  theme_void() +
  theme(plot.caption = element_text(size = 12)) +
  labs(caption = bquote(lambda == .(sf.bir$area[1] / sf.bir_circ$area[1])))


p.bir_rect <- ggplot() +
  geom_sf(data = sf.bir_rect, fill = "white") +
  geom_sf(data = sf.bir, fill = "red") +
  theme_void() +
  theme(plot.caption = element_text(size = 12)) +
  labs(caption = bquote(lambda == .(sf.bir$area[1] / sf.bir_rect$area[1])))

p.bir_circ + p.bir_rect
ggsave("figures/bir_buffers.png", plot = p.bir_circ + p.bir_rect)
```
Chunk for the creation of summary statistics

```{r}
df.top_100 <- read_csv("data/fragmentation_top_100.csv")

top10_rect <- df.top_100 %>% arrange(-index_rect) %>% slice(1:10)
bottom10_rect <- df.top_100 %>% arrange(index_rect) %>% slice(1:10)
top10_circ <- df.top_100 %>% arrange(-index_circ) %>% slice(1:10)
bottom10_circ <- df.top_100 %>% arrange(index_circ) %>% slice(1:10)
combined_rect <- bind_rows(top10_rect, bottom10_rect) %>% 
  select(cbsatitle,index_rect,index_rect_central,num_munis,pop_muni,hhi_pop_muni)
combined_circ <- bind_rows(top10_circ, bottom10_circ) %>% 
  select(cbsatitle,index_circ,index_circ_central,num_munis,pop_muni,hhi_pop_muni)

combined_rect %>%
  kable("latex", 
        booktabs = TRUE, 
        linesep = "", 
        caption = "Top and Bottom 10 MSAs by Spatial Fragmentation (Bounding Rectangle)",
        col.names = c("CBSA Title","SDI","SDI - Central","Num. Munis","Muni Pop.","HHI - Muni Pop.")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(1:20, extra_latex_after = "\\addlinespace[0.2em]") %>% 
  row_spec(10, hline_after = TRUE) %>% 
  pack_rows("Top 10", 1, 10, latex_gap_space = "0.7em") %>%
  pack_rows("Bottom 10", 11, 20, latex_gap_space = "0.7em") %>%
  save_kable("figures/spat_frag_sumstats_rect.tex")

combined_circ %>%
  kable("latex", 
        booktabs = TRUE, 
        linesep = "", 
        caption = "Top and Bottom 10 MSAs by Spatial Fragmentation (Bounding Circle)",
        col.names = c("CBSA Title","SDI","SDI - Central","Num. Munis","Muni Pop.","HHI - Muni Pop.")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(1:20, extra_latex_after = "\\addlinespace[0.2em]") %>% 
  row_spec(10, hline_after = TRUE) %>% 
  pack_rows("Top 10", 1, 10, latex_gap_space = "0.7em") %>%
  pack_rows("Bottom 10", 11, 20, latex_gap_space = "0.7em") %>%
  save_kable("figures/spat_frag_sumstats_circ.tex")
```

