---
title: "Agglomeration - ZBP"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 0, digits = 6)  
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, plotly, tidycensus, readxl, httr, jsonlite, sf, mapview, tigris, lehdr, spatstat, crsuggest, dbmss, stargazer) 
options(tigris_use_cache = TRUE)
# census_api_key("", install = TRUE)
setwd("~/Github/honors-thesis/[04] Agglomeration")
```

## Data Assembly

### Crosswalk Creation

Initialize by creating a set of crosswalk files to ensure ZIP-CBSA-County compatibility

```{r}
# use constant cbsa-county pairs over time

xwalk.cbsa_fips <- read_csv("https://data.nber.org/cbsa-csa-fips-county-crosswalk/2023/cbsa2fipsxw_2023.csv") %>% 
  mutate(fullfips = paste0(fipsstatecode, fipscountycode))

xwalk.states <- fips_codes

hud_key <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiOTkyMTgyZTJhMjMzYWIxYjg4ZTg0ZWNhNzc4YTgxZmUyZjBiMjI2YjBjN2I1Nzc2MDExZWE0NjQyMWYyMGVjMjM4M2U5MzVjOTQwNDQ2MGQiLCJpYXQiOjE3NjI1Mzg0NTEuMTcxNzA0LCJuYmYiOjE3NjI1Mzg0NTEuMTcxNzM4LCJleHAiOjIwNzgwNzEyNTEuMTU2ODU1LCJzdWIiOiIxMTI2MDMiLCJzY29wZXMiOltdfQ.VUVEmcDRysXhP5os4E3v9J4HrcFU10lycZShAQszZBUwyujCIMoFK_5WbKbKMRvvN9GJ_-mF3eJ_QcnIGuLGWg"
hud_url <- "https://www.huduser.gov/hudapi/public/usps"

# use type = 7 for county -> zip, 8 for cbsa -> zip

response <- httr::GET(hud_url, 
                      query = list(type = 2, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
zip_county <- as.data.frame(df)

response <- httr::GET(hud_url, 
                      query = list(type = 3, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
cbsa_zip <- as.data.frame(df)
rm(df, json_data, response)

# extract appropriate columns from both xwalks. preserve business ratio

xwalk.zip_county <- zip_county %>% 
  rename(zip = data.results.zip,
         county = data.results.geoid,
         bus_ratio = data.results.bus_ratio) %>% 
  select(county, zip, bus_ratio)

xwalk.zip_cbsa <- cbsa_zip %>% 
  rename(cbsa = data.results.geoid,
         zip = data.results.zip,
         bus_ratio = data.results.bus_ratio) %>% 
  select(cbsa, zip, bus_ratio)

# check the distribution: most are 0 or 100

#ggplot() + 
  #geom_histogram(data = xwalk.zip_county, aes(x = bus_ratio))
```

### ZBP Data Pull

Now pull County Business Patterns data. I use 2019 data for easier compatibility with other Census products. Pulling from an API tends to be slower than the manual download, which is available here: https://www.census.gov/data/datasets/2019/econ/cbp/2019-cbp.html, though I include a function to download as well.

```{r}
get_zbp_detail <- function(year) {
  base <- "https://www2.census.gov/programs-surveys/cbp/datasets"
  url  <- sprintf("%s/%s/zbp%02ddetail.zip", base, year, year %% 100)

  tmp_zip <- tempfile(fileext = ".zip")
  tmp_dir <- tempfile()

  download.file(url, tmp_zip, mode = "wb")
  unzip(tmp_zip, exdir = tmp_dir)

  txt_file <- list.files(tmp_dir, pattern = "\\.txt$", full.names = TRUE)
  df <- read_csv(txt_file, show_col_types = FALSE)

  # clean up
  unlink(tmp_zip)
  unlink(tmp_dir, recursive = TRUE)

  df
}

zbp_2019 <- get_zbp_detail(2019)

# other option
# zbp.2019 <- read_csv("[04] Agglomeration/raw data/zbp19detail.txt")

names(zbp_2019)

#zbp_inspect <- zbp_23 %>% 
#  filter(zip == "01001") %>% 
#  filter(grepl("-", naics) == FALSE) %>% 
#  filter(grepl("/", naics) == FALSE) %>% 
#  summarise(count = sum(est))

zbp_2019_long <- zbp_2019 %>% 
  select(-c(city, stabbr, cty_name)) %>% 
  mutate(across(`n<5`:`n1000`, ~ replace_na(as.numeric(.x), 0))) %>% 
  pivot_longer(
    cols = `est`:`n1000`,
    names_to = "est_desc",
    values_to = "est"
  ) %>% 
  inner_join(xwalk.zip_cbsa, by = join_by(zip == zip), relationship = "many-to-many") %>% 
  rename(bus_ratio_cbsa = bus_ratio) %>% 
  mutate(adj_est = est * bus_ratio_cbsa)

# relationships will be many to many since there is occasionally overlap for zip codes between counties/cbsas. above code is set for cbsas, but can also look at specific counties

#zbp_2019_long_county <- zbp_2019_long %>% 
#  inner_join(xwalk.zip_county, by = join_by(zip == zip)) %>%
#  rename(bus_ratio_county = bus_ratio)

summary(zbp_2019_long)

# Additionally, load a ZIP sf
geo.zcta <- zctas(
  cb = TRUE,
  year = 2019
)
```

### Sanity Check: Scientific Research

To ensure the ZBP file is working properly, I perform a quick following chunk, pulling out data for specific industries from the ZBP data. This first run will use "Scientific Research and Development Services" as an example.

```{r}
# Filter for NAICS codes beginning with 5417. 
zbp_2019_research <- zbp_2019_long %>% 
  filter(naics == "5417//",
         est_desc == "est")

# Get a sf file as well
geo.zcta_research <- geo.zcta %>% 
  left_join(zbp_2019_research, by = join_by(ZCTA5CE10 == zip)) %>% 
  filter(!is.na(est))

# mapview(geo.zcta.research) # View the complete range of data

# Check the pull for a specific CBSA to ensure it's fully functional. One can also adjust the filtered city to provide a mapview for CBSAs that may interest them.

geo.zcta_spec <- geo.zcta_research %>% 
  filter(cbsa == "14460")

mapview(geo.zcta_spec, zcol = "est")
```

## Constructing metrics

### ZBP HHI

The first ZBP metric will be a firm-based HHI, which is much less computationally complex than the Ripley L calculation from LODES blocks. This will give a rough approximation of how concentrated industry-level activity is across the US. I focus on the high-tech industry, using BDS-HT NAICS codes given by [Hecker (2024)](https://www.federalreserve.gov/econres/notes/feds-notes/high-tech-business-entry-in-the-pandemic-era-20240419.html).

```{r}
# input codes here 
naics_codes = c(5112,5182,5417,5415) %>% 
  as.character() %>% 
  str_pad(6, "right", "/")

# get adjusted employment by bus_ratio

naics_to_hhi <- function(naics_vec) {
  codes_filtered <- zbp_2019_long %>% 
    filter(naics %in% naics_vec,
           est_desc == "est")
  
  get_hhi <- function(est_col) {
  total = sum(est_col)
  shares = est_col/total
  result = sum(shares^2)
  result
  }
  
  codes_filtered <- codes_filtered %>% 
    group_by(cbsa) %>% 
    summarise(hhi = get_hhi(adj_est),
              total_est = sum(adj_est)) %>% 
    filter(total_est > 50)
}

zbp_2019_hhi <- naics_to_hhi(naics_codes)

ggplot(data = zbp_2019_hhi, aes(x = log(total_est), y = hhi)) +
  geom_point()
```

A clear pattern emerges from these HHIs: they are strongly decreasing in the total number of establishments within a CBSA. Intuitively, this makes sense, as there are likely just not enough establishments in smaller CBSAs to make the HHI meaningful.

While this means the measure may not be capturing sufficient variance between similarly sized CBSAs, I perform a basic set of regressions to check feasibility.

```{r}
# Load fragmentation data
setwd("~/Github/honors-thesis/[03] Fragmentation")
df.frag = read_csv("data/fragmentation_basic.csv") %>% 
  mutate(cbsacode = as.character(cbsacode))

df.reg <- inner_join(
  df.frag, zbp_2019_hhi, by = join_by(cbsacode == cbsa)
)

reg.model1 <- lm(hhi ~ num_munis, data = df.reg)
reg.model2 <- lm(hhi ~ num_munis + pop_muni, data = df.reg)
stargazer(reg.model1, reg.model2, type = "text")
ggplot(data = df.reg, aes(x = num_munis, y = hhi)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Controlling for total municipal population, the number of municipalities in a CBSA has a negative effect on firm clustering at the 10% level. I note that this is the regression in its most basic form: identification issues discussed in the final paper preclude these figures from approaching a true causal estimate.

### LODES Blocks

Now, I move to a more precise measure of firm clustering. The procedure below will be performed for the Seattle-Tacoma-Bellevue metropolitan area. First, one directly grabs the LODES data. I also load in a preparatory ZBP file filtered by the needed industries.

```{r}
df.lodes <- grab_lodes(
  "wa",
  2019,
  version = "LODES7",
  lodes_type = "wac",
  job_type = "JT00",
  segment = "S000",
  agg_geo = "block",
  use_cache = TRUE
)

zbp_2019_ind <- zbp_2019_long %>% 
  filter(naics %in% naics_codes,
         # cbsa == "42660",
         est_desc == "est")

# Construct baseline for manufacturing
naics_codes_baseline <- c("31----","32----","33----")

zbp_2019_baseline <- zbp_2019_long %>% 
  filter(naics %in% naics_codes_baseline,
         est_desc == "est")

n_unique(zbp_2019_ind$naics)

# Quick check to visualize
#mapview(geo.zcta %>% 
#          inner_join(zbp_2019_ind, by = join_by(ZCTA5CE10 == zip)) %>% 
#          group_by(ZCTA5CE10) %>% 
#          summarise(total_est = sum(est)),
#        zcol = "total_est")
```

Perform statistical analysis on the enriched LODES points. Reference LODES variables documentation [here](https://lehd.ces.census.gov/data/lodes/LODES8/LODESTechDoc8.0.pdf), showing that CNS09 (corresponding to NAICS 51) and CNS12 (corresponding to NAICS 54) cover the needed NAICS codes.
To find the number of relevant establishments per block, I perform a matching procedure where potential point locations are weighted within ZIP codes by their share of LODES employment, then assigned firms based on these probabilities.

```{r}
# Function to fetch blocks as points given a CBSA code
get_block_points <- function(cbsacode) {
  x <- xwalk.cbsa_fips %>%
    filter(cbsacode == {{ cbsacode }}) %>%
    select(fipsstatecode, fipscountycode) %>%
    distinct()

  counties_by_state <- split(x$fipscountycode, x$fipsstatecode)

  result <- purrr::imap(counties_by_state, ~ blocks(
      state = .y,
      county = .x,
      year = 2010,
      progress_bar = TRUE
    )) %>%
    bind_rows() %>%
    st_centroid()
  
  # Use projected geometry
  crs <- suggest_top_crs(result)
  result <- result %>% 
    st_transform(crs)
  
  return(result)
}

geo.block_points <- get_block_points(42660)
crs <- suggest_top_crs(geo.block_points)


# May need to think more about how the function will be constructed(NAICS beforehand or called within?)

point_match <- function(block_points) {
  # Filter points where there is nonzero employment in both CNS09 and CNS12
  needed_crs <- suggest_top_crs(block_points)

  needed_sts <- 
    xwalk.states %>% 
    filter(state_code %in% unique(block_points$STATEFP)) %>% 
    distinct(state) %>% pull(state) %>% 
    tolower()
  
  lodes <- grab_lodes(
  needed_sts,
  2019,
  version = "LODES7",
  lodes_type = "wac",
  job_type = "JT00",
  segment = "S000",
  agg_geo = "block",
  use_cache = TRUE
  )
  
  # This gives all needed points in the CBSA with employment
 block_ind <- block_points %>% 
  st_transform(needed_crs) %>% 
  select(GEOID10) %>% 
  inner_join(lodes, by = join_by(GEOID10 == w_geocode)) %>% 
  select(GEOID10, starts_with("CNS"))
  }

geo.block_ind <- point_match(geo.block_points)

block_zip_join <- function(block_ind, zbp_df) {
  zcta <- geo.zcta %>% 
  inner_join(zbp_df, by = join_by(ZCTA5CE10 == zip)) %>% 
  select(-GEOID10) %>% 
  st_transform(crs)
  
  full <- st_join(block_ind, zcta, join = st_within) %>% 
  filter(!is.na(ZCTA5CE10))
}

geo.ind_full <- block_zip_join(geo.block_ind, zbp_2019_ind)
geo.base_full <- block_zip_join(geo.block_ind, zbp_2019_baseline)

# Perform the matching procedure
geo.block_est <- geo.ind_full %>% 
  select(GEOID10, CNS09, CNS12, ZCTA = ZCTA5CE10, naics, adj_est) %>%
  mutate(x = st_coordinates(geo.ind_full)[, 1],
         y = st_coordinates(geo.ind_full)[, 2]) %>%
  st_drop_geometry() %>% 
  group_by(ZCTA) %>% 
  mutate(CNS09_total = sum(CNS09) / n_unique(naics),
         CNS12_total = sum(CNS12) / n_unique(naics),
         ) %>% 
  ungroup() %>% 
  mutate(CNS09_pct = CNS09 /CNS09_total,
         CNS12_pct = CNS12 / CNS12_total,
         est_point = case_when(
           naics == "5112//" ~ adj_est * CNS09_pct,
           naics == "5182//" ~ adj_est * CNS09_pct,
           naics == "5415//" ~ adj_est * CNS12_pct,
           naics == "5417//" ~ adj_est * CNS12_pct
         )) %>% 
  group_by(GEOID10) %>% 
  summarise(est_total = sum(est_point),
            x = first(x),
            y = first(y)) %>% 
  filter(est_total > 0) # remove zeros

# Also perform for baseline. Function call may be needed in future, though would have to proof a crosswalk between the CNS and NAICS codes

geo.block_est_base <- geo.base_full %>% 
  select(GEOID10, CNS05, ZCTA = ZCTA5CE10, naics, adj_est) %>%
  mutate(x = st_coordinates(geo.base_full)[, 1],
         y = st_coordinates(geo.base_full)[, 2]) %>%
  st_drop_geometry() %>% 
  group_by(ZCTA) %>% 
  mutate(CNS05_total = sum(CNS05) / n_unique(naics)) %>% 
  ungroup() %>% 
  mutate(CNS05_pct = CNS05 /CNS05_total,
         est_point = case_when(
           naics == "31----" ~ adj_est * CNS05_pct,
           naics == "32----" ~ adj_est * CNS05_pct,
           naics == "33----" ~ adj_est * CNS05_pct,
         )) %>% 
  group_by(GEOID10) %>% 
  summarise(est_total = sum(est_point),
            x = first(x),
            y = first(y)) %>% 
  filter(est_total > 0)

# Check to make sure sums still correspond with the actual number of establishments in ZBP data
sum(geo.block_est$est_total)
sum(geo.block_est_base$est_total)

zbp_2019_long %>% filter(cbsa == "42660",
                         est_desc == "est",
                         naics %in% naics_codes) %>% 
  summarise(n = sum(est)) %>% 
  pull(n)

zbp_2019_baseline %>% filter(cbsa == "42660",
                         est_desc == "est",
                         naics %in% naics_codes_baseline) %>% 
  summarise(n = sum(est)) %>% 
  pull(n)
  

ggplot(geo.block_est, aes(x=x,y=y)) +
  geom_point()

# Optional view
# mapview(geo.test %>% filter(est_total > 0.25), zcol = "est_total")
```
### K-Density

Before performing the main spatial statistics, let's get a simpler LODES aggregated dataset to test for clustering at broad NAICS levels. This assumes that points have already been collected from the previous chunk.

```{r}
crs <- suggest_top_crs(geo.block_points)
geo.block_lodes <- geo.block_points %>% 
  st_transform(crs) %>% 
  select(GEOID10) %>% 
  inner_join(df.lodes, by = join_by(GEOID10 == w_geocode)) %>% 
  mutate(x = st_coordinates(geometry)[,1],
         y = st_coordinates(geometry)[,2]) %>%
  st_drop_geometry() %>% 
  select(GEOID10,x,y,starts_with("CNS")) %>% 
  pivot_longer(cols = starts_with("CNS"),
               names_to = "type",
               values_to = "weights") %>% 
  filter(weights != 0)

wmppp.lodes <- wmppp(
  df = data.frame(
    X = geo.block_lodes$x,
    Y = geo.block_lodes$y,
    PointType = geo.block_lodes$type,
    PointWeight = geo.block_lodes$weights
  ),
  window = owin(
      xrange = range(geo.block_lodes$x),
      yrange = range(geo.block_lodes$y)
  )
)

kd_lodes <- map_dfr(c("CNS05","CNS07","CNS10","CNS12"), function(rt) {
  as.data.frame(Kdhat(
  X = wmppp.lodes,
  ReferenceType = rt,
  NeighborType = rt,
  Weighted = TRUE
)) %>% 
    mutate(type = rt)
})

p.kd_lodes <- ggplot(data = kd_lodes, aes(x=r/1000,y=Kd)) +
  geom_line() + 
  facet_wrap(~type) +
  theme_bw() +
  ylab("K-Density") + xlab("Distance (km)")

p.kd_lodes

ggsave("figures/kd_lodes.png", plot=p.kd_lodes)
```

Now perform for the selected NAICS codes with baseline manufacturing calculated

```{r}
A_points <- geo.block_est %>% 
  mutate(est_total = round(0.15+est_total))

baseline_points <- geo.block_est_base %>% 
  mutate(est_total = round(0.15+est_total))

expanded_wmppp <- function(points, point_type) {
  points_expanded <- points[rep(seq_len(nrow(points)), points$est_total), ] %>% 
    mutate(est_total = 1,
           PointType = point_type) %>% 
    rename(PointWeight = est_total) %>% 
    select(-GEOID10)
  
  points_wmppp <- wmppp(
    df = points_expanded,
    window = owin(
      xrange = range(points_expanded$x),
      yrange = range(points_expanded$y)
    )
  )
  
  result_wmppp <- rjitter(points_wmppp, radius = 50, retry = TRUE)
  return(result_wmppp)
}

wmppp.A_points <- expanded_wmppp(A_points, "A")
wmppp.baseline <- expanded_wmppp(baseline_points, "baseline")
wmppp.full <- superimpose(wmppp.A_points, wmppp.baseline)

kd.A <- map_dfr(c("A","baseline"), function(rt) {
  as.data.frame(Kdhat(
  X = wmppp.full,
  ReferenceType = rt,
  NeighborType = rt,
  Weighted = TRUE
)) %>% 
    mutate(type = rt)
})

plan(multisession, workers = availableCores() - 1)

kdenv.A <- KdEnvelope(
  X = wmppp.full,
  ReferenceType = "A",
  NumberOfSimulations = 100,
  Alpha = 0.05,
  Global = TRUE,
  parallel = TRUE
)
```
Plotting figures from above

```{r}
ggplot(data = kd.A, aes(x=r/1000,y=Kd,col=type)) +
  geom_line()

p.kd_ind <- ggplot(data = kdenv.A) + 
  geom_line(aes(x=r/1000,y=obs)) +
  geom_line(aes(x=r/1000,y=hi), linetype = "dashed") +
  geom_line(aes(x=r/1000,y=lo), linetype = "dashed") +
  theme_bw() +
  ylab("K-Density") + xlab("Distance (km)")

ggsave("figures/kd_ind.png", plot=p.kd_ind)
```

## Figures

```{r}
plot_zbp <- zbp_2019_long %>% 
  filter(cbsa == "42660",
         naics %in% naics_codes) %>% 
  group_by(zip) %>% 
  summarise(est = sum(est))

p.zbp <- ggplot(data = geo.zcta %>% 
         inner_join(plot_zbp, by = join_by(ZCTA5CE10 == zip)),
       aes(fill = est)) +
  geom_sf() +
  scale_fill_gradient(low = "blue", high = "yellow") +
  theme_void() +
  labs(fill = "Adjusted \nEstablishments") 

ggsave("figures/seattle_zbp.png", plot = p.zbp)


plot_points <- geo.block_est %>% 
  st_as_sf(coords = c("x","y"), crs = 6596) %>% 
  st_intersection(places(state = "WA", cb = TRUE) %>% 
  filter(NAME == "Seattle") %>% 
    st_transform(6596))

p.lodes <- ggplot(data = plot_points, aes(color = est_total)) + # Use color, not fill
  geom_sf(size = 0.7) + 
  scale_color_gradient(low = "blue", high = "yellow") + # Use scale_color
  theme_void() +
  labs(color = "Adjusted \nEstablishments")

p.lodes

ggsave("figures/seattle_lodes.png", plot = p.lodes)
```

## Testing Space

Testing block sizes for inputs into rjitter. Find that median block size is around 50 meters, which seems like a reasonable input given that high tech establishments cluster in more dense urban areas with smaller block sizes.

```{r}
test_blocks <- blocks(
  state = "WA",
  county = "King"
) %>% 
  mutate(area = st_area(geometry),
         rad = sqrt(area)/pi)

median(test_blocks$rad)
```

Testing error induced by rounding for creating establishment points.

```{r}
# Actual establishments
zbp_2019_long %>% filter(cbsa == "42660",
                         est_desc == "est",
                         naics %in% naics_codes) %>% 
  summarise(n = sum(est)) %>% 
  pull(n)

# Current estimated
sum(geo.block_est$est_total)

# Choose parameter that approximates the previous sum
test_points <- geo.block_est %>% 
  mutate(est_total = round(0.17 + est_total))

sum(test_points$est_total)

# Construct new points 
test_points_expanded <- test_points[rep(seq_len(nrow(test_points)), test_points$est_total), ] %>% 
  mutate(est_total = 1,
         PointType = "A") %>% 
  rename(PointWeight = est_total) %>% 
  select(-GEOID10)

test_wmppp <- wmppp(
  df = test_points_expanded,
  window = owin(
      xrange = range(test_points_expanded$x),
      yrange = range(test_points_expanded$y)
  )
)

test_wmppp_2 <- rjitter(test_wmppp, radius = 50, retry = TRUE)

test_kd <- Kdhat(test_wmppp_2,
                 ReferenceType = "A")

ggplot(data = test_kd, aes(x=r/1000,y=Kd)) +
  geom_line()
```

