---
title: "agglomeration_lightweight"
output: html_document
date: "2026-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 0, digits = 6)  
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, plotly, tidycensus, readxl, httr, jsonlite, sf, mapview, tigris, lehdr, spatstat, crsuggest, dbmss, stargazer, data.table, stringi) 
options(tigris_use_cache = TRUE)
```

## Set up

Initialize with complete set of crosswalk files (copied from agglomeration_main.Rmd). Use consistent nomenclature: 'cbsa', 'statefips', 'countyfips', 'fips', 'state_abbr', 'state_name'

```{r}
xwalk.cbsa_fips <- read_csv("https://data.nber.org/cbsa-csa-fips-county-crosswalk/2023/cbsa2fipsxw_2023.csv") %>% 
  mutate(fips = paste0(fipsstatecode, fipscountycode)) %>% 
  select(cbsa = cbsacode, statefips = fipsstatecode, countyfips = fipscountycode, fips)

xwalk.state_fips <- fips_codes %>% 
  select(state_abbr = state, state_name, statefips = state_code, county_name = county, countyfips = county_code)

hud_key <- read_file("hud_key.txt")
hud_url <- "https://www.huduser.gov/hudapi/public/usps"

response <- httr::GET(hud_url, 
                      query = list(type = 2, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
zip_county <- as.data.frame(df)

response <- httr::GET(hud_url, 
                      query = list(type = 3, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
cbsa_zip <- as.data.frame(df)
rm(df, json_data, response)

# extract appropriate columns from both xwalks. preserve business ratio

xwalk.zip_fips <- zip_county %>% 
  rename(zip = data.results.zip,
         county = data.results.geoid,
         bus_ratio = data.results.bus_ratio) %>% 
  select(fips = county, zip, bus_ratio)
rm(zip_county)

xwalk.zip_cbsa <- cbsa_zip %>% 
  rename(cbsa = data.results.geoid,
         zip = data.results.zip,
         bus_ratio = data.results.bus_ratio) %>% 
  select(cbsa, zip, bus_ratio)
rm(cbsa_zip)
```

Get national block and LODES data, join for nation-wide df.

```{r}
# storage folder
folder <- "C:/Users/phynm/OneDrive/Documents/thesis_data/"

blk <- fread(file.path(folder, "blocks_2020.csv"))
setDT(blk)
blk[, fips := sprintf("%05.0f", GEOID20 / 1e10)]
names(blk)
setnames(blk, 
         old = c("GEOID20","INTPTLON20","INTPTLAT20"), 
         new = c("geoid","lon","lat"))

lodes <- fread(file.path(folder, "lodes_2020.csv"))
setDT(lodes)
setnames(lodes,
         old = "w_geocode", new = "geoid")

blk <- blk %>% 
  inner_join(lodes, by = "geoid")
```

Find and clean ZBP data with CBSA matching

```{r}
zbp <- fread(file.path(folder, "zbp20detail.txt")) %>% 
  select(-c(city, stabbr, cty_name)) %>% 
  mutate(across(`n<5`:`n1000`, ~ replace_na(as.numeric(.x), 0)),
         zip = sprintf("%05d", zip)) %>% 
  pivot_longer(
    cols = `est`:`n1000`,
    names_to = "est_desc",
    values_to = "est"
  ) %>% 
  inner_join(xwalk.zip_cbsa, by = join_by(zip == zip), relationship = "many-to-many") %>% 
  rename(bus_ratio_cbsa = bus_ratio) %>% 
  mutate(adj_est = est * bus_ratio_cbsa) %>% 
  filter(est_desc == "est") %>% 
  select(-est_desc, -est, -bus_ratio_cbsa)

setDT(zbp)

# get a zcta sf as well
sf.zcta <- zctas(
  cb = TRUE,
  year = 2020
)

# st_crs(sf.zcta)
```

Get a ZCTA match for each point (can take a bit, necessary for matching)

```{r}
sf.zcta <- zctas(
  cb = TRUE,
  year = 2020
) %>% 
  select(zip = ZCTA5CE20)

sf.blk <- st_as_sf(blk %>% 
                     select(geoid, lon, lat),
                   coords = c("lon", "lat"), crs = 4269)

blk_zcta <- st_join(sf.blk, sf.zcta, join = st_within)
xwalk.geoid_zip <- blk_zcta %>% 
  st_drop_geometry() %>% 
  select(geoid, zip)
```



## Function

Get the needed LODES data for the naics codes

```{r}
n_naics <- c(711130)

format_naics <- function(codes) {
  a <- as.character(codes)

  b <- ifelse(
    nchar(a) <= 2,
    str_pad(a, width = 6, side = "right", pad = "-"),
    str_pad(a, width = 6, side = "right", pad = "/")
  )
  
  b

  return(append(b, "31----"))
}

selected_lodes <- function(f_naics) {
  
xwalk.cns_naics <- fread("data/cns_naics_xwalk.csv") %>% 
  mutate(naics = as.character(naics))
  
f_cns <- xwalk.cns_naics %>% 
  filter(naics %in% (substr(format_naics(f_naics), start = 1, stop = 2) %>% 
           unique())
  ) %>% 
  pull(cns) %>% 
  unique()

f_lodes <- blk %>% 
  select(geoid, all_of(f_cns))
setDT(f_lodes)

return(f_lodes)
}


naics_to_cns <- function(f_naics) {
  xwalk.cns_naics <- fread("data/cns_naics_xwalk.csv") %>% 
    mutate(naics = as.character(naics))
  
f_cns <- xwalk.cns_naics %>% 
  filter(naics %in% (substr(format_naics(f_naics), start = 1, stop = 2) %>% 
           unique())
  ) %>% 
  pull(cns) %>% 
  unique() 

return(f_cns)
}

r_lodes <- selected_lodes(n_naics) %>% 
  inner_join(xwalk.geoid_zip, by = "geoid")

r_zbp <- zbp %>% 
  filter(naics %in% format_naics(n_naics)) %>% 
  select(-name) %>% 
  pivot_wider(id_cols = c("zip","cbsa"),
              names_prefix = "naics",
              names_from = naics,
              values_from = adj_est)

r_full <- inner_join(
  r_lodes, r_zbp,
  by = "zip",
  relationship = "many-to-many"
) %>% 
  filter(cbsa != "99999") # no rural areas

enrich_points <- function(f_naics) {

f_lodes <- selected_lodes(f_naics)

f_zbp <- zbp %>% 
  filter(naics %in% format_naics(f_naics)) %>% 
  select(-name)


}
```

Create large LODES df 

```{r}
match_establishments <- function(full_df, n_naics) {
  
  naics_cols <- paste0("naics", format_naics(n_naics))
  cns_cols   <- naics_to_cns(n_naics)
  
  purrr::map2_dfr(naics_cols, cns_cols, function(naics_col, cns_col) {
    
    full_df %>%
      filter(
        !is.na(.data[[naics_col]]), .data[[naics_col]] > 0,
        !is.na(.data[[cns_col]]),   .data[[cns_col]]  > 0
      ) %>%
      group_by(zip) %>%
      mutate(
        weight      = .data[[cns_col]] / sum(.data[[cns_col]], na.rm = TRUE),
        est_assigned = .data[[naics_col]] * weight   # fractional establishments per block
      ) %>%
      ungroup() %>%
      mutate(naics = naics_col, cns = cns_col)
  })
}

t <- match_establishments(r_full, n_naics)

v <- inner_join(
  sf.blk,
  t %>% 
    filter(cbsa == "42660"),
  by = "geoid"
)

ggplot(v) +
  geom_sf()
```



## Appendix

Create a df with every TIGER block as indexed by central point.

```{r}
folder <- "C:/Users/phynm/OneDrive/Documents/thesis_data/"

states <- state.abb

get_block_coords <- function(st) {
  blocks <- blocks(state = st, year = 2020)
  st_drop_geometry(blocks) %>% 
    select(GEOID20, INTPTLON20, INTPTLAT20)
}

us_block_coords <- map_df(states, get_block_coords)

write_csv(us_block_coords, file.path(folder, "blocks_2020.csv"))
```

Fetch LODES data
```{r}
lodes <- function(st) {
  grab_lodes(
    st,
    2020,
    version = "LODES8",
    lodes_type = "wac",
    job_type = "JT00",
    segment = "S000",
    agg_geo = "block",
    use_cache = TRUE
  )
}

us_block_lodes <- map_df(states, possibly(
  lodes,
  otherwise = NULL,
  quiet = FALSE)
)

names(us_block_lodes)
simp <- us_block_lodes %>% 
  select(w_geocode, starts_with("CNS"))

write_csv(simp, file.path(folder, "lodes_2020.csv"))
```

```{r}
y <- zbp$naics[substr(zbp$naics, 1, 2) == "32"] %>% 
  unique()

y
```

