---
title: "Agglomeration - ZBP"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 0, digits = 6)  
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, plotly, tidycensus, readxl, httr, jsonlite, sf, mapview, tigris, lehdr, spatstat, crsuggest, dbmss, stargazer) 
setwd("~/Github/honors-thesis/[04] Agglomeration")
```

## Data Assembly

### Crosswalk Creation

Initialize by creating a set of crosswalk files to ensure ZIP-CBSA-County compatibility

```{r}
# use constant cbsa-county pairs over time

xwalk.cbsa_fips <- read_csv("https://data.nber.org/cbsa-csa-fips-county-crosswalk/2023/cbsa2fipsxw_2023.csv") %>% 
  mutate(fullfips = paste0(fipsstatecode, fipscountycode))

hud_key <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiOTkyMTgyZTJhMjMzYWIxYjg4ZTg0ZWNhNzc4YTgxZmUyZjBiMjI2YjBjN2I1Nzc2MDExZWE0NjQyMWYyMGVjMjM4M2U5MzVjOTQwNDQ2MGQiLCJpYXQiOjE3NjI1Mzg0NTEuMTcxNzA0LCJuYmYiOjE3NjI1Mzg0NTEuMTcxNzM4LCJleHAiOjIwNzgwNzEyNTEuMTU2ODU1LCJzdWIiOiIxMTI2MDMiLCJzY29wZXMiOltdfQ.VUVEmcDRysXhP5os4E3v9J4HrcFU10lycZShAQszZBUwyujCIMoFK_5WbKbKMRvvN9GJ_-mF3eJ_QcnIGuLGWg"
hud_url <- "https://www.huduser.gov/hudapi/public/usps"

# use type = 7 for county -> zip, 8 for cbsa -> zip

response <- httr::GET(hud_url, 
                      query = list(type = 2, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
zip_county <- as.data.frame(df)

response <- httr::GET(hud_url, 
                      query = list(type = 3, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
cbsa_zip <- as.data.frame(df)
rm(df, json_data, response)

# extract appropriate columns from both xwalks. preserve business ratio

xwalk.zip_county <- zip_county %>% 
  rename(zip = data.results.zip,
         county = data.results.geoid,
         bus_ratio = data.results.bus_ratio) %>% 
  select(county, zip, bus_ratio)

xwalk.zip_cbsa <- cbsa_zip %>% 
  rename(cbsa = data.results.geoid,
         zip = data.results.zip,
         bus_ratio = data.results.bus_ratio) %>% 
  select(cbsa, zip, bus_ratio)

# check the distribution: most are 0 or 100

#ggplot() + 
  #geom_histogram(data = xwalk.zip_county, aes(x = bus_ratio))
```

### ZBP Data Pull

Now pull County Business Patterns data. I use 2019 data for easier compatibility with other Census products. Pulling from an API tends to be slower than the manual download, which is available here: https://www.census.gov/data/datasets/2019/econ/cbp/2019-cbp.html, though I include a function to download as well.

```{r}
get_zbp_detail <- function(year) {
  base <- "https://www2.census.gov/programs-surveys/cbp/datasets"
  url  <- sprintf("%s/%s/zbp%02ddetail.zip", base, year, year %% 100)

  tmp_zip <- tempfile(fileext = ".zip")
  tmp_dir <- tempfile()

  download.file(url, tmp_zip, mode = "wb")
  unzip(tmp_zip, exdir = tmp_dir)

  txt_file <- list.files(tmp_dir, pattern = "\\.txt$", full.names = TRUE)
  df <- read_csv(txt_file, show_col_types = FALSE)

  # clean up
  unlink(tmp_zip)
  unlink(tmp_dir, recursive = TRUE)

  df
}

zbp_2019 <- get_zbp_detail(2019)

# other option
# zbp.2019 <- read_csv("[04] Agglomeration/raw data/zbp19detail.txt")

names(zbp_2019)

#zbp_inspect <- zbp_23 %>% 
#  filter(zip == "01001") %>% 
#  filter(grepl("-", naics) == FALSE) %>% 
#  filter(grepl("/", naics) == FALSE) %>% 
#  summarise(count = sum(est))

zbp_2019_wide <- zbp_2019 %>% 
  select(-c(city, stabbr, cty_name)) %>% 
  mutate(across(`n<5`:`n1000`, ~ replace_na(as.numeric(.x), 0))) %>% 
  pivot_longer(
    cols = `est`:`n1000`,
    names_to = "emp_desc",
    values_to = "emp"
  ) %>% 
  inner_join(xwalk.zip_cbsa, by = join_by(zip == zip)) %>% 
  rename(bus_ratio_cbsa = bus_ratio)

# relationships will be many to many since there is occasionally overlap for zip codes between counties/cbsas. above code is set for cbsas, but can also look at specific counties

#zbp_2019_wide_county <- zbp_2019_wide %>% 
#  inner_join(xwalk.zip_county, by = join_by(zip == zip)) %>%
#  rename(bus_ratio_county = bus_ratio)

summary(zbp_2019_wide)
```

Now we can pull out data for specific industries from the ZBP data. This first run will use "Scientific Research and Development Services" as an example.

```{r}
# NAICS codes beginning with 5417. 

zbp_2019_research <- zbp_2019_wide %>% 
  filter(naics == "5417//",
         emp_desc == "est")

geo.zcta <- zctas(
  cb = TRUE,
  year = 2019,
)

geo.zcta_research <- geo.zcta %>% 
  left_join(zbp_2019_research, by = join_by(ZCTA5CE10 == zip)) %>% 
  filter(!is.na(emp))

# mapview(geo.zcta.research)
```

Check the pull for a specific CBSA to ensure it's fully functional. One can also adjust the filtered city to provide a mapview for CBSAs that may interest them.

```{r}
geo.zcta_spec <- geo.zcta_research %>% 
  filter(cbsa == "14460")

mapview(geo.zcta_spec, zcol = "emp")
```

## Constructing metrics

### ZBP HHI

The first ZBP metric will be a firm-based HHI, which is much less computationally complex than the Ripley L calculation from LODES blocks. This will give a rough approximation of how concentrated industry-level activity is across the US.

```{r}
# input codes here 
naics_codes = c(5112,5161,5181,5182,5415) %>% 
  as.character() %>% 
  str_pad(6, "right", "/")

# get adjusted employment by bus_ratio

naics_to_hhi <- function(naics_vec) {
  code_filter <- zbp_2019_wide %>% 
    filter(naics %in% naics_vec,
           emp_desc == "est") %>% 
    mutate(adj_emp = emp * bus_ratio_cbsa)
  
  get_hhi <- function(emp_col) {
  total = sum(emp_col)
  shares = emp_col/total
  result = sum(shares^2)
  result
  }
  
  code_filter <- code_filter %>% 
    group_by(cbsa) %>% 
    summarise(hhi = get_hhi(adj_emp),
              total_emp = sum(adj_emp)) %>% 
    filter(total_emp > 100)
}

zbp_2019_ind <- naics_to_hhi(naics_codes)

ggplot(data = zbp_2019_ind, aes(x = total_emp, y = hhi)) +
  geom_point()
```

```{r}
setwd("~/Github/honors-thesis/[03] Fragmentation")
df.frag = read_csv("data/fragmentation_basic.csv") %>% 
  mutate(cbsacode = as.character(cbsacode))

df.reg <- inner_join(
  df.frag, zbp_2019_ind, by = join_by(cbsacode == cbsa)
)

reg.model1 <- lm(hhi ~ num_munis, data = df.reg)
reg.model2 <- lm(hhi ~ num_munis + pop_muni, data = df.reg)
stargazer(reg.model1, reg.model2, type = "text")
ggplot(data = df.reg, aes(x = num_munis, y = hhi)) +
  geom_point()
```



### LODES Blocks

Directly grab the LODES data

```{r}
df.lodes <- grab_lodes(
  "wa",
  2019,
  version = "LODES7",
  lodes_type = "wac",
  job_type = "JT00",
  segment = "S000",
  agg_geo = "block",
  use_cache = TRUE
)
```

Perform statistical analysis on the enriched LODES points

```{r}
get_block_points <- function(cbsacode) {
  x <- xwalk.cbsa_fips %>%
    filter(cbsacode == {{ cbsacode }}) %>%
    select(fipsstatecode, fipscountycode) %>%
    distinct()

  counties_by_state <- split(x$fipscountycode, x$fipsstatecode)

  purrr::imap(counties_by_state, ~ blocks(
      state = .y,
      county = .x,
      year = 2010,
      progress_bar = FALSE
    )) %>%
    bind_rows() %>%
    st_centroid()
  }

geo.block_points <- get_block_points(42660)

# use projected geometry
crs <- suggest_top_crs(geo.block_points)
geo.block_points <- geo.block_points %>% 
  st_transform(crs) %>% 
  mutate(x = st_coordinates(geo.block_points)[,1],
         y = st_coordinates(geo.block_points)[,2]) %>% 
  select(GEOID10,x,y) %>% 
  st_drop_geometry() %>% 
  inner_join(df.lodes, by = join_by(GEOID10 == w_geocode))

my_wmppp <- wmppp(
  data.frame(
    X           = geo.block_points$x,
    Y           = geo.block_points$y,
    PointType   = "Block",
    PointWeight = geo.block_points$CNS12
  ),
  window = owin(
    xrange = range(geo.block_points$x),
    yrange = range(geo.block_points$y)
  )
)

result <- Lest(my_wmppp, correction = "Ripley")

ggplot() +
  geom_line(data = result, aes(x = r, y = iso))
```

