---
title: "agglomeration_lightweight"
output: html_document
date: "2026-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 10, digits = 6)  
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, plotly, tidycensus, readxl, httr, jsonlite, sf, mapview, tigris, lehdr, spatstat, crsuggest, dbmss, stargazer, data.table, stringi, future) 
options(tigris_use_cache = TRUE)
```

## Set up

Set the needed year:

```{r}
n_year <- 2020
```


Initialize with complete set of crosswalk files (copied from agglomeration_main.Rmd). Use consistent nomenclature: 'cbsa', 'statefips', 'countyfips', 'fips', 'state_abbr', 'state_name'

```{r}
xwalk.cbsa_fips <- read_csv("https://data.nber.org/cbsa-csa-fips-county-crosswalk/2023/cbsa2fipsxw_2023.csv") %>% 
  mutate(fips = paste0(fipsstatecode, fipscountycode)) %>% 
  select(cbsa = cbsacode, statefips = fipsstatecode, countyfips = fipscountycode, fips)

xwalk.state_fips <- fips_codes %>% 
  select(state_abbr = state, state_name, statefips = state_code, county_name = county, countyfips = county_code)

hud_key <- read_file("hud_key.txt")
hud_url <- "https://www.huduser.gov/hudapi/public/usps"

response <- httr::GET(hud_url, 
                      query = list(type = 2, query = "ALL", year = n_year),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
zip_county <- as.data.frame(df)

response <- httr::GET(hud_url, 
                      query = list(type = 3, query = "ALL", year = n_year),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
cbsa_zip <- as.data.frame(df)
rm(df, json_data, response)

# extract appropriate columns from both xwalks. preserve business ratio

xwalk.zip_fips <- zip_county %>% 
  rename(zip = data.results.zip,
         county = data.results.geoid,
         bus_ratio = data.results.bus_ratio) %>% 
  select(fips = county, zip, bus_ratio)
rm(zip_county)

xwalk.zip_cbsa <- cbsa_zip %>% 
  rename(cbsa = data.results.geoid,
         zip = data.results.zip,
         bus_ratio = data.results.bus_ratio) %>% 
  select(cbsa, zip, bus_ratio)
rm(cbsa_zip)
```

Get national block and LODES data, join for nation-wide df.

```{r}
# storage folder
folder <- "C:/Users/phynm/OneDrive/Documents/thesis_data/"

blk <- fread(file.path(folder, "blocks_2020.csv"))
setDT(blk)
blk[, fips := sprintf("%05.0f", GEOID20 / 1e10)]
names(blk)
setnames(blk, 
         old = c("GEOID20","INTPTLON20","INTPTLAT20"), 
         new = c("geoid","lon","lat"))

lodes <- fread(file.path(folder, "lodes_2020.csv"))
setDT(lodes)
setnames(lodes,
         old = "w_geocode", new = "geoid")

blk <- blk %>% 
  inner_join(lodes, by = "geoid")

rm(lodes)
```

Find and clean ZBP data with CBSA matching

```{r}
zbp <- fread(file.path(folder, "zbp20detail.txt")) %>% 
  select(-c(city, stabbr, cty_name)) %>% 
  mutate(across(`n<5`:`n1000`, ~ replace_na(as.numeric(.x), 0)),
         zip = sprintf("%05d", zip)) %>% 
  pivot_longer(
    cols = `est`:`n1000`,
    names_to = "est_desc",
    values_to = "est"
  ) %>% 
  inner_join(xwalk.zip_cbsa, by = join_by(zip == zip), relationship = "many-to-many") %>% 
  rename(bus_ratio_cbsa = bus_ratio) %>% 
  mutate(adj_est = est * bus_ratio_cbsa) %>% 
  filter(est_desc == "est") %>% 
  select(-est_desc, -est, -bus_ratio_cbsa)

setDT(zbp)
```

Get a ZCTA match for each point (can take a bit, necessary for matching)

```{r}
sf.zcta <- zctas(
  cb = TRUE,
  year = n_year
) %>% 
  select(zip = ZCTA5CE20)

sf.blk <- st_as_sf(blk %>% 
                     select(geoid, lon, lat),
                   coords = c("lon", "lat"), crs = 4269)

blk_zcta <- st_join(sf.blk, sf.zcta, join = st_within)
xwalk.geoid_zip <- blk_zcta %>% 
  st_drop_geometry() %>% 
  select(geoid, zip)
```



## Function

Set up all the needed helper functions for the creation of the final dataset

```{r}
# Format raw NAICS codes to be compatible with ZBP data
format_naics <- function(codes) {
  a <- as.character(codes)

  b <- ifelse(
    nchar(a) <= 2,
    str_pad(a, width = 6, side = "right", pad = "-"),
    str_pad(a, width = 6, side = "right", pad = "/")
  )
  
  b

  return(append(b, "31----"))
}

# Select needed cols from LODES data based on NAICS
selected_lodes <- function(f_naics) {
  
xwalk.cns_naics <- fread("data/cns_naics_xwalk.csv") %>% 
  mutate(naics = as.character(naics))
  
f_cns <- xwalk.cns_naics %>% 
  filter(naics %in% (substr(format_naics(f_naics), start = 1, stop = 2) %>% 
           unique())
  ) %>% 
  pull(cns) %>% 
  unique()

f_lodes <- blk %>% 
  select(geoid, all_of(f_cns))
setDT(f_lodes)

return(f_lodes)
}

# Similar to above, but simply return list of needed CNS by NAICS code
naics_to_cns <- function(f_naics) {
  xwalk.cns_naics <- fread("data/cns_naics_xwalk.csv") %>% 
    mutate(naics = as.character(naics))
  
  a <- format_naics(f_naics)
  
  map_chr(a, function(n) {
    prefix <- substr(n, 1, 2)
    xwalk.cns_naics %>%
      filter(naics == prefix) %>%
      pull(cns) %>%
      .[[1]]
  })
}

# Join the selected points with ZCTAs (crosswalk created above)
enrich_points <- function(f_naics) {

f_lodes <- selected_lodes(f_naics) %>% 
  inner_join(xwalk.geoid_zip, by = "geoid")

f_zbp <- zbp %>% 
  filter(naics %in% format_naics(f_naics)) %>% 
  select(-name) %>% 
  pivot_wider(id_cols = c("zip","cbsa"),
              names_prefix = "naics",
              names_from = naics,
              values_from = adj_est)

f_full <- inner_join(
  f_lodes, f_zbp,
  by = "zip",
  relationship = "many-to-many"
) %>% 
  filter(cbsa != "99999") # no rural areas

return(f_full)
}
```

Primary matching function

```{r}
match_establishments <- function(f_naics) {
  
  full_df <- enrich_points(f_naics)
  
  naics_cols <- paste0("naics", format_naics(f_naics))
  cns_cols   <- naics_to_cns(f_naics)
  
  map2_dfr(naics_cols, cns_cols, function(naics_col, cns_col) {
    
    full_df %>%
      filter(
        !is.na(.data[[naics_col]]), .data[[naics_col]] > 0,
        !is.na(.data[[cns_col]]),   .data[[cns_col]]  > 0
      ) %>%
      group_by(zip) %>%
      mutate(
        weight      = .data[[cns_col]] / sum(.data[[cns_col]], na.rm = TRUE),
        est_assigned = .data[[naics_col]] * weight   # fractional establishments per block
      ) %>%
      ungroup() %>%
      mutate(naics = sub("^naics", "", naics_col), cns = cns_col)
  }) %>% 
    select(geoid, est_assigned, naics, cbsa, zip)
}
```

Distribute points to integers instead of fractions

```{r}
largest_remainder <- function(x) {
  floored    <- floor(x)
  remainder  <- x - floored
  shortage   <- round(sum(x)) - sum(floored)
  floored + (rank(-remainder, ties.method = "first") <= shortage)
}

format_match <- function(df) {
  f_v <- df %>%
  group_by(zip, naics) %>%
  mutate(est_assigned_int = largest_remainder(est_assigned)) %>%
  ungroup() %>% 
  mutate(type =
           ifelse(
             naics == "31----",
             "baseline", "A"
           )
  )
  
  f_w <- inner_join(
  sf.blk, f_v, by = "geoid"
)
  
  return(f_w)
}
```

KD estimating function

```{r}
expanded_wmppp <- function(points, point_type, est_col = "est_assigned_int") {
  points[rep(seq_len(nrow(points)), points[[est_col]]), ] %>%
    mutate(est_total = 1,
           PointType  = point_type) %>%
    rename(PointWeight = est_total) %>%
    wmppp(
      df     = .,
      window = owin(xrange = range(.$x), yrange = range(.$y))
    ) %>%
    rjitter(radius = 50, retry = TRUE)
  }

kd_main <- function(match_df, f_cbsa) {
  
f_cbsa <- as.character(f_cbsa)

f_epsg <- cbsa_proj %>% 
    filter(cbsa == f_cbsa) %>% 
    pull(epsg)

f_cbsa_df <- w %>%
    filter(cbsa == f_cbsa) %>%
    st_transform(f_epsg)

f_coords <- st_coordinates(f_cbsa_df)
  
f_cbsa_df <- f_cbsa_df %>%
    st_drop_geometry() %>%
    mutate(x = f_coords[, "X"], y = f_coords[, "Y"])

f_wmppp_baseline <- f_cbsa_df %>% 
  filter(type == "baseline") %>% 
  expanded_wmppp("baseline")

f_wmppp_a        <- f_cbsa_df %>% 
  filter(type == "A")        %>% 
  expanded_wmppp("A")

f_wmppp_full <- superimpose(f_wmppp_a, f_wmppp_baseline)

f_kdenv <- KdEnvelope(
    X                 = f_wmppp_full,
    r                 = c(0, 250, 500, 1000),
    ReferenceType     = "A",
    NumberOfSimulations = 100,
    Alpha             = 0.05,
    Global            = TRUE,
    parallel          = FALSE
  )

# plan(sequential)
  

return(f_kdenv)
}

# id fetch
get_ids <- function(threshold, f_naics) {

  threshold <- as.numeric(threshold)
  
cbsa_ids <- zbp %>% 
  filter(naics %in% format_naics(f_naics),
         cbsa != "99999") %>% 
  group_by(cbsa) %>% 
  summarise(n = sum(adj_est)) %>% 
  arrange(-n) %>% 
  filter(n >= threshold) %>% 
  pull(cbsa)

return(cbsa_ids)
}
```

## Main call

I separate out the different parts of the function for easier checks against errors.

```{r}
t0 <- Sys.time()

# Enter needed NAICS codes
n_naics <- c(71)

x <- match_establishments(n_naics)
y <- format_match(x)

# Choose the threshold for establishments (set at 250 for CBSA)
i <- get_ids(250, n_naics)

r_kd <- map(i, function(id) {
  message("Processing CBSA: ", id)
  tryCatch(
    kd_main(y, f_cbsa = id),
    error = function(e) {
      warning("CBSA ", id, " failed: ", e$message)
      NULL
    }
  )
}) %>%
  setNames(i) %>%
  compact()

r_final <- map_dfr(names(r_kd), ~ 
  as.data.frame(r_kd[[.x]]) %>% mutate(cbsa = .x)
) %>% 
  mutate(gamma = pmax(obs - hi, 0))

t1 <- Sys.time()
elapsed <- t1 - t0
elapsed

write_csv(r_final, "data/kd_creative.csv")
```

## Data visualization

```{r}
t <- comb %>% 
  filter(r != 0) %>% 
  arrange(-gamma) %>% 
  slice(1:20)
t

u <- w %>% 
  filter(
    cbsa == "14460",
    type == "A",
    est_assigned_int != 0
  )

mapview(u, zcol = "est_assigned_int")
```


## Appendix

Create a df with every TIGER block as indexed by central point.

```{r}
folder <- "C:/Users/phynm/OneDrive/Documents/thesis_data/"

states <- state.abb

get_block_coords <- function(st) {
  blocks <- blocks(state = st, year = 2020)
  st_drop_geometry(blocks) %>% 
    select(GEOID20, INTPTLON20, INTPTLAT20)
}

us_block_coords <- map_df(states, get_block_coords)

write_csv(us_block_coords, file.path(folder, "blocks_2020.csv"))
```

Fetch LODES data

```{r}
lodes <- function(st) {
  grab_lodes(
    st,
    2020,
    version = "LODES8",
    lodes_type = "wac",
    job_type = "JT00",
    segment = "S000",
    agg_geo = "block",
    use_cache = TRUE
  )
}

us_block_lodes <- map_df(states, possibly(
  lodes,
  otherwise = NULL,
  quiet = FALSE)
)

names(us_block_lodes)
simp <- us_block_lodes %>% 
  select(w_geocode, starts_with("CNS"))

write_csv(simp, file.path(folder, "lodes_2020.csv"))
```

Create a CBSA-UTM crosswalk to project points. First, create CBSA sf

```{r}
sf.cbsa <- counties(
  state = NULL,
  cb = TRUE,
  year = n_year
) %>% 
  left_join(xwalk.cbsa_fips, by = join_by(GEOID == fips)) %>% 
  select(cbsa) %>% 
  filter(!is.na(cbsa)) %>% 
  group_by(cbsa) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_as_sf()

lon_to_utm_epsg <- function(lon) {
  zone <- floor((lon + 180) / 6) + 1
  32600 + zone
}

# Get one centroid per CBSA to determine its UTM zone
cbsa_proj <- sf.cbsa %>% 
  mutate(
    centroid = st_centroid(geometry),
    lon      = st_coordinates(centroid)[, "X"],
    epsg     = lon_to_utm_epsg(lon)
  ) %>%
  st_drop_geometry() %>%
  select(cbsa, epsg)
```

ZBP NAICS look up

```{r}
all_codes <- zbp %>% 
  select(naics) %>% 
  unique() %>% 
  pull(naics)

all_codes %>% 
  filter(starts_with("71"))
```


Old version of the main call

```{r}
autoplot(z)

cbsa_ids <- zbp %>% 
  filter(naics %in% format_naics(n_naics),
         cbsa != "99999") %>% 
  group_by(cbsa) %>% 
  summarise(n = sum(adj_est)) %>% 
  arrange(-n) %>% 
  filter(n >= 100) %>% 
  slice(1:100) %>% 
  pull(cbsa)

cbsa_ids2 <- zbp %>% 
  filter(naics %in% format_naics(n_naics),
         cbsa != "99999") %>% 
  group_by(cbsa) %>% 
  summarise(n = sum(adj_est)) %>% 
  arrange(-n) %>% 
  filter(n >= 250) %>% 
  slice(101:207) %>% 
  pull(cbsa) 

results2 <- map(cbsa_ids2, function(id) {
  message("Processing CBSA: ", id)
  tryCatch(
    kd_main(w, f_cbsa = id),
    error = function(e) {
      warning("CBSA ", id, " failed: ", e$message)
      NULL
    }
  )
}) %>%
  setNames(cbsa_ids2) %>%
  compact()

comb <- map_dfr(names(results), ~ 
  as.data.frame(results[[.x]]) %>% mutate(cbsa = .x)
) %>% 
  mutate(gamma = pmax(obs - hi, 0))

comb2 <- map_dfr(names(results2), ~ 
  as.data.frame(results2[[.x]]) %>% mutate(cbsa = .x)
) %>% 
  mutate(gamma = pmax(obs - hi, 0))

c <- bind_rows(comb, comb2)

write_csv(c, "data/kd_tech.csv")
```


