---
title: "Agglomeration - ZBP"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 0, digits = 6)  
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, plotly, tidycensus, readxl, httr, jsonlite, sf, mapview, tigris, lehdr, spatstat, crsuggest, dbmss, stargazer) 
options(tigris_use_cache = TRUE)
setwd("~/Github/honors-thesis/[04] Agglomeration")
```

## Data Assembly

### Crosswalk Creation

Initialize by creating a set of crosswalk files to ensure ZIP-CBSA-County compatibility

```{r}
# use constant cbsa-county pairs over time

xwalk.cbsa_fips <- read_csv("https://data.nber.org/cbsa-csa-fips-county-crosswalk/2023/cbsa2fipsxw_2023.csv") %>% 
  mutate(fullfips = paste0(fipsstatecode, fipscountycode))

hud_key <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiOTkyMTgyZTJhMjMzYWIxYjg4ZTg0ZWNhNzc4YTgxZmUyZjBiMjI2YjBjN2I1Nzc2MDExZWE0NjQyMWYyMGVjMjM4M2U5MzVjOTQwNDQ2MGQiLCJpYXQiOjE3NjI1Mzg0NTEuMTcxNzA0LCJuYmYiOjE3NjI1Mzg0NTEuMTcxNzM4LCJleHAiOjIwNzgwNzEyNTEuMTU2ODU1LCJzdWIiOiIxMTI2MDMiLCJzY29wZXMiOltdfQ.VUVEmcDRysXhP5os4E3v9J4HrcFU10lycZShAQszZBUwyujCIMoFK_5WbKbKMRvvN9GJ_-mF3eJ_QcnIGuLGWg"
hud_url <- "https://www.huduser.gov/hudapi/public/usps"

# use type = 7 for county -> zip, 8 for cbsa -> zip

response <- httr::GET(hud_url, 
                      query = list(type = 2, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
zip_county <- as.data.frame(df)

response <- httr::GET(hud_url, 
                      query = list(type = 3, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
cbsa_zip <- as.data.frame(df)
rm(df, json_data, response)

# extract appropriate columns from both xwalks. preserve business ratio

xwalk.zip_county <- zip_county %>% 
  rename(zip = data.results.zip,
         county = data.results.geoid,
         bus_ratio = data.results.bus_ratio) %>% 
  select(county, zip, bus_ratio)

xwalk.zip_cbsa <- cbsa_zip %>% 
  rename(cbsa = data.results.geoid,
         zip = data.results.zip,
         bus_ratio = data.results.bus_ratio) %>% 
  select(cbsa, zip, bus_ratio)

# check the distribution: most are 0 or 100

#ggplot() + 
  #geom_histogram(data = xwalk.zip_county, aes(x = bus_ratio))
```

### ZBP Data Pull

Now pull County Business Patterns data. I use 2019 data for easier compatibility with other Census products. Pulling from an API tends to be slower than the manual download, which is available here: https://www.census.gov/data/datasets/2019/econ/cbp/2019-cbp.html, though I include a function to download as well.

```{r}
get_zbp_detail <- function(year) {
  base <- "https://www2.census.gov/programs-surveys/cbp/datasets"
  url  <- sprintf("%s/%s/zbp%02ddetail.zip", base, year, year %% 100)

  tmp_zip <- tempfile(fileext = ".zip")
  tmp_dir <- tempfile()

  download.file(url, tmp_zip, mode = "wb")
  unzip(tmp_zip, exdir = tmp_dir)

  txt_file <- list.files(tmp_dir, pattern = "\\.txt$", full.names = TRUE)
  df <- read_csv(txt_file, show_col_types = FALSE)

  # clean up
  unlink(tmp_zip)
  unlink(tmp_dir, recursive = TRUE)

  df
}

zbp_2019 <- get_zbp_detail(2019)

# other option
# zbp.2019 <- read_csv("[04] Agglomeration/raw data/zbp19detail.txt")

names(zbp_2019)

#zbp_inspect <- zbp_23 %>% 
#  filter(zip == "01001") %>% 
#  filter(grepl("-", naics) == FALSE) %>% 
#  filter(grepl("/", naics) == FALSE) %>% 
#  summarise(count = sum(est))

zbp_2019_long <- zbp_2019 %>% 
  select(-c(city, stabbr, cty_name)) %>% 
  mutate(across(`n<5`:`n1000`, ~ replace_na(as.numeric(.x), 0))) %>% 
  pivot_longer(
    cols = `est`:`n1000`,
    names_to = "est_desc",
    values_to = "est"
  ) %>% 
  inner_join(xwalk.zip_cbsa, by = join_by(zip == zip), relationship = "many-to-many") %>% 
  rename(bus_ratio_cbsa = bus_ratio) %>% 
  mutate(adj_est = est * bus_ratio_cbsa)

# relationships will be many to many since there is occasionally overlap for zip codes between counties/cbsas. above code is set for cbsas, but can also look at specific counties

#zbp_2019_long_county <- zbp_2019_long %>% 
#  inner_join(xwalk.zip_county, by = join_by(zip == zip)) %>%
#  rename(bus_ratio_county = bus_ratio)

summary(zbp_2019_long)

# Additionally, load a ZIP sf
geo.zcta <- zctas(
  cb = TRUE,
  year = 2019
)
```

### Sanity Check: Scientific Research

To ensure the ZBP file is working properly, I perform a quick following chunk, pulling out data for specific industries from the ZBP data. This first run will use "Scientific Research and Development Services" as an example.

```{r}
# Filter for NAICS codes beginning with 5417. 
zbp_2019_research <- zbp_2019_long %>% 
  filter(naics == "5417//",
         est_desc == "est")

# Get a sf file as well
geo.zcta_research <- geo.zcta %>% 
  left_join(zbp_2019_research, by = join_by(ZCTA5CE10 == zip)) %>% 
  filter(!is.na(est))

# mapview(geo.zcta.research) # View the complete range of data

# Check the pull for a specific CBSA to ensure it's fully functional. One can also adjust the filtered city to provide a mapview for CBSAs that may interest them.

geo.zcta_spec <- geo.zcta_research %>% 
  filter(cbsa == "14460")

mapview(geo.zcta_spec, zcol = "est")
```

## Constructing metrics

### ZBP HHI

The first ZBP metric will be a firm-based HHI, which is much less computationally complex than the Ripley L calculation from LODES blocks. This will give a rough approximation of how concentrated industry-level activity is across the US. I focus on the high-tech industry, using BDS-HT NAICS codes given by [Hecker (2024)](https://www.federalreserve.gov/econres/notes/feds-notes/high-tech-business-entry-in-the-pandemic-era-20240419.html).

```{r}
# input codes here 
naics_codes = c(5112,5182,5417,5415) %>% 
  as.character() %>% 
  str_pad(6, "right", "/")

# get adjusted employment by bus_ratio

naics_to_hhi <- function(naics_vec) {
  codes_filtered <- zbp_2019_long %>% 
    filter(naics %in% naics_vec,
           est_desc == "est")
  
  get_hhi <- function(est_col) {
  total = sum(est_col)
  shares = est_col/total
  result = sum(shares^2)
  result
  }
  
  codes_filtered <- codes_filtered %>% 
    group_by(cbsa) %>% 
    summarise(hhi = get_hhi(adj_est),
              total_est = sum(adj_est)) %>% 
    filter(total_est > 50)
}

zbp_2019_hhi <- naics_to_hhi(naics_codes)

ggplot(data = zbp_2019_hhi, aes(x = log(total_est), y = hhi)) +
  geom_point()
```
A clear pattern emerges from these HHIs: they are strongly decreasing in the total number of establishments within a CBSA. Intuitively, this makes sense, as there are likely just not enough establishments in smaller CBSAs to make the HHI meaningful.

While this means the measure may not be capturing sufficient variance between similarly sized CBSAs, I perform a basic set of regressions to check feasibility.

```{r}
# Load fragmentation data
setwd("~/Github/honors-thesis/[03] Fragmentation")
df.frag = read_csv("data/fragmentation_basic.csv") %>% 
  mutate(cbsacode = as.character(cbsacode))

df.reg <- inner_join(
  df.frag, zbp_2019_hhi, by = join_by(cbsacode == cbsa)
)

reg.model1 <- lm(hhi ~ num_munis, data = df.reg)
reg.model2 <- lm(hhi ~ num_munis + pop_muni, data = df.reg)
stargazer(reg.model1, reg.model2, type = "text")
ggplot(data = df.reg, aes(x = num_munis, y = hhi)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Controlling for total municipal population, the number of municipalities in a CBSA has a negative effect on firm clustering at the 10% level. I note that this is the regression in its most basic form: identification issues discussed in the final paper preclude these figures from approaching a true causal estimate.

### LODES Blocks

Now, I move to a more precise measure of firm clustering. The procedure below will be performed for the Seattle-Tacoma-Bellevue metropolitan area. First, one directly grabs the LODES data. I also load in a preparatory ZBP file filtered by the needed industries.

```{r}
df.lodes <- grab_lodes(
  "wa",
  2019,
  version = "LODES7",
  lodes_type = "wac",
  job_type = "JT00",
  segment = "S000",
  agg_geo = "block",
  use_cache = TRUE
)

zbp_2019_ind <- zbp_2019_long %>% 
  filter(naics %in% naics_codes,
         # cbsa == "42660",
         est_desc == "est")

n_unique(zbp_2019_ind$naics)

# Quick check to visualize
#mapview(geo.zcta %>% 
#          inner_join(zbp_2019_ind, by = join_by(ZCTA5CE10 == zip)) %>% 
#          group_by(ZCTA5CE10) %>% 
#          summarise(total_est = sum(est)),
#        zcol = "total_est")
```

Perform statistical analysis on the enriched LODES points. Reference LODES variables documentation [here](https://lehd.ces.census.gov/data/lodes/LODES8/LODESTechDoc8.0.pdf), showing that CNS09 (corresponding to NAICS 51) and CNS12 (corresponding to NAICS 54) cover the needed NAICS codes.
To find the number of relevant establishments per block, I perform a matching procedure where potential point locations are weighted within ZIP codes by their share of LODES employment, then assigned firms based on these probabilities.

```{r}
# Function to fetch blocks as points given a CBSA code
get_block_points <- function(cbsacode) {
  x <- xwalk.cbsa_fips %>%
    filter(cbsacode == {{ cbsacode }}) %>%
    select(fipsstatecode, fipscountycode) %>%
    distinct()

  counties_by_state <- split(x$fipscountycode, x$fipsstatecode)

  result <- purrr::imap(counties_by_state, ~ blocks(
      state = .y,
      county = .x,
      year = 2010,
      progress_bar = TRUE
    )) %>%
    bind_rows() %>%
    st_centroid()
  
  # Use projected geometry
  crs <- suggest_top_crs(result)
  result <- result %>% 
    st_transform(crs)
  
  return(result)
}

geo.block_points <- get_block_points(42660) 

## reassemble this function. will need to create a new LODES fetch for multiple states like above

point_match <- function(block_points) {
# Filter points where there is nonzero employment in both CNS09 and CNS12
geo.block_ind <- geo.block_points %>% 
  st_transform(crs) %>% 
  select(GEOID10) %>% 
  inner_join(df.lodes, by = join_by(GEOID10 == w_geocode)) %>% 
  select(GEOID10, CNS09, CNS12) %>% 
  filter(CNS09 > 0 | CNS12 > 0)

# Create sf ZIP code file
geo.zcta_ind <- geo.zcta %>% 
  inner_join(zbp_2019_ind, by = join_by(ZCTA5CE10 == zip)) %>% 
  select(-GEOID10) %>% 
  st_transform(crs)

# Block-ZIP full joined 
geo.ind_full <- st_join(geo.block_ind, geo.zcta_ind, join = st_within) %>% 
  filter(!is.na(ZCTA5CE10))

# Perform the matching procedure
geo.block_est <- geo.ind_full %>% 
  select(GEOID10, CNS09, CNS12, ZCTA = ZCTA5CE10, naics, adj_est) %>%
  mutate(x = st_coordinates(geo.ind_full)[,1],
         y = st_coordinates(geo.ind_full)[,2]) %>%
  st_drop_geometry() %>% 
  group_by(ZCTA) %>% 
  mutate(CNS09_total = sum(CNS09) / n_unique(naics),
         CNS12_total = sum(CNS12) / n_unique(naics),
         ) %>% 
  ungroup() %>% 
  mutate(CNS09_pct = CNS09 /CNS09_total,
         CNS12_pct = CNS12 / CNS12_total,
         est_point = case_when(
           naics == "5112//" ~ adj_est * CNS09_pct,
           naics == "5182//" ~ adj_est * CNS09_pct,
           naics == "5415//" ~ adj_est * CNS12_pct,
           naics == "5417//" ~ adj_est * CNS12_pct
         )) %>% 
  group_by(GEOID10) %>% 
  summarise(est_total = sum(est_point),
            x = first(x),
            y = first(y)) %>% 
  filter(est_total > 0) # remove zeros

return(geo.block_est)
}

geo.block_est <- point_match(geo.block_points)

# Check to make sure sums still correspond with the actual number of establishments in ZBP data
sum(geo.block_est$est_total)

zbp_2019_long %>% filter(cbsa == "42660",
                         est_desc == "est",
                         naics %in% naics_codes) %>% 
  summarise(n = sum(est)) %>% 
  pull(n)

# Optional view
# mapview(geo.test %>% filter(est_total > 0.25), zcol = "est_total")
```

Perform spatial statistics with the assembled point-establishment dataset.

```{r}
test_wmppp <- wmppp(
  data.frame(
    X           = geo.block_est$x,
    Y           = geo.block_est$y,
    PointType   = "Block",
    PointWeight = geo.block_est$est_total
  ),
  window = owin(
    xrange = range(geo.block_est$x),
    yrange = range(geo.block_est$y)
  )
) %>% 
  rescale(s = 1000, unitname = "km")

plot(test_wmppp)

result <- Lest(test_wmppp, correction = "Ripley", rmax = 100)

ggplot() +
  geom_line(data = result, aes(x = r, y = iso)) +
  geom_line(data = result, aes(x = r, y = theo))
```

