---
title: "THESIS WORKFILE"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 0, digits = 3)  
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, plotly, tidycensus, readxl, httr, jsonlite, sf, mapview, tigris, lehdr) 
```

## ZBP Assembly

### Crosswalk Creation

Initialize by creating a set of crosswalk files to ensure ZIP-CBSA-County compatibility

```{r}
# use constant cbsa-county pairs over time

xwalk.cbsa_fips <- read_csv("https://data.nber.org/cbsa-csa-fips-county-crosswalk/2023/cbsa2fipsxw_2023.csv") %>% 
  mutate(fullfips = paste0(fipsstatecode, fipscountycode))

hud_key <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiOTkyMTgyZTJhMjMzYWIxYjg4ZTg0ZWNhNzc4YTgxZmUyZjBiMjI2YjBjN2I1Nzc2MDExZWE0NjQyMWYyMGVjMjM4M2U5MzVjOTQwNDQ2MGQiLCJpYXQiOjE3NjI1Mzg0NTEuMTcxNzA0LCJuYmYiOjE3NjI1Mzg0NTEuMTcxNzM4LCJleHAiOjIwNzgwNzEyNTEuMTU2ODU1LCJzdWIiOiIxMTI2MDMiLCJzY29wZXMiOltdfQ.VUVEmcDRysXhP5os4E3v9J4HrcFU10lycZShAQszZBUwyujCIMoFK_5WbKbKMRvvN9GJ_-mF3eJ_QcnIGuLGWg"
hud_url <- "https://www.huduser.gov/hudapi/public/usps"

# use type = 7 for county -> zip, 8 for cbsa -> zip

response <- httr::GET(hud_url, 
                      query = list(type = 2, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
zip_county <- as.data.frame(df)

response <- httr::GET(hud_url, 
                      query = list(type = 3, query = "ALL", year = 2017),
                      add_headers(Authorization = paste("Bearer", hud_key))
)

stop_for_status(response)
json_data <- httr::content(response, as = "text", encoding = "UTF-8")
df <- jsonlite::fromJSON(json_data)
cbsa_zip <- as.data.frame(df)
rm(df, json_data, response)

# extract appropriate columns from both xwalks. preserve business ratio

xwalk.zip_county <- zip_county %>% 
  rename(zip = data.results.zip,
         county = data.results.geoid,
         bus_ratio = data.results.bus_ratio) %>% 
  select(county, zip, bus_ratio)

xwalk.zip_cbsa <- cbsa_zip %>% 
  rename(cbsa = data.results.geoid,
         zip = data.results.zip,
         bus_ratio = data.results.bus_ratio) %>% 
  select(cbsa, zip, bus_ratio)

# check the distribution: most are 0 or 100

#ggplot() + 
  #geom_histogram(data = xwalk.zip_county, aes(x = bus_ratio))
```

### Data Pull

Now pull County Business Patterns data. I use 2019 data for easier compatibility with other Census products. Pulling from an API tends to be slower than the manual download, which is available here: https://www.census.gov/data/datasets/2019/econ/cbp/2019-cbp.html, though I include a function to download as well.

```{r}
get_zbp_detail <- function(year) {
  base <- "https://www2.census.gov/programs-surveys/cbp/datasets"
  url  <- sprintf("%s/%s/zbp%02ddetail.zip", base, year, year %% 100)

  tmp_zip <- tempfile(fileext = ".zip")
  tmp_dir <- tempfile()

  download.file(url, tmp_zip, mode = "wb")
  unzip(tmp_zip, exdir = tmp_dir)

  txt_file <- list.files(tmp_dir, pattern = "\\.txt$", full.names = TRUE)

  df <- read_csv(txt_file, show_col_types = FALSE)

  # Clean up
  unlink(tmp_zip)
  unlink(tmp_dir, recursive = TRUE)

  df
}

zbp_2019 <- get_zbp_detail(2019)

# other option
# zbp.2019 <- read_csv("[04] Agglomeration/raw data/zbp19detail.txt")

names(zbp_2019)

#zbp_inspect <- zbp_23 %>% 
#  filter(zip == "01001") %>% 
#  filter(grepl("-", naics) == FALSE) %>% 
#  filter(grepl("/", naics) == FALSE) %>% 
#  summarise(count = sum(est))

zbp_2019_wide <- zbp_2019 %>% 
  select(-c(city, stabbr, cty_name)) %>% 
  mutate(across(`n<5`:`n1000`, ~ replace_na(as.numeric(.x), 0))) %>% 
  pivot_longer(
    cols = `est`:`n1000`,
    names_to = "emp_desc",
    values_to = "emp"
  ) %>% 
  inner_join(xwalk.zip_county, by = join_by(zip == zip)) %>% 
  arrange(bus_ratio)

test <- zbp_2019_wide %>% 
  filter(bus_ratio != 1) %>% 
  filter(bus_ratio != 0) %>% 
  arrange(zip, naics)

summary(zbp_2019_wide)
```

Now we can pull out data for specific industries from the ZBP data. This first run will use "Scientific Research and Development Services" as an example.

```{r}
# NAICS codes beginning with 5417. 

zbp.2019.research <- zbp.2019.wide %>% 
  filter(naics == "5417//",
         emp_desc == "est")

geo.zcta <- zctas(
  cb = TRUE,
  year = 2019,
)

geo.zcta.research <- geo.zcta %>% 
  left_join(zbp.2019.research, by = join_by(ZCTA5CE10 == zip)) %>% 
  filter(!is.na(emp))

# mapview(geo.zcta.research)